
Libraries Used
```{r}
library(zipcodeR)
library(mice)
library(tidyverse)
```


Import CSV Dataset
```{r}
data <- read.csv('/Users/gabby/Documents/Class Archive/Spring 2023/Senior Project/Senior_Stats_data.csv',
                 header = TRUE,
                 na.strings = "", ",")


```


## Changing Datatypes
-----------------------

converting certain variables from char to numeric
```{r}
data$efc <- as.numeric(data$efc)
data$term_gpa <- as.numeric(data$term_gpa)
data$act_score <- as.numeric(data$act_score)
data$total_hrs_earned <- as.numeric(data$total_hrs_earned)
data$transfer_gpa <- as.numeric(data$transfer_gpa)
data$hs_gpa <- as.numeric(data$hs_gpa)
```

Converting all characters to factors
This allows for easier access to contents in each variable. This is especially
important because this data set has so many observations.
```{r}
data[sapply(data, is.character)] <- lapply(data[sapply(data, is.character)], 
                                       as.factor)
```


## Imputation
--------------------------------------------------------------------------

Dealing with NA/ missing/ 'UNKOWN' values
These are values that are technically missing, but not inputed as NA, which
is how R recognized a value is missing. The value may be 'Unknown' instead of NA
for example. This is a problem because the imputation function that I will 
employ only imputes NA values.

variables w/missing values that are not inputed as NA
```{r} 
summary(data$first_gen_status)
summary(data$sex)
summary(data$dependency_status)
summary(data$has_children)
```

CONVERTING 'unk' factors to NA to prepare for imputation
```{r}
data <- set_na(data, na = c(first_gen_status = 'Unknown',
               sex = 'N',
               dependency_status = 'unk',
               has_children = 'unk'))
  
```



Creates a new dataframe 'missing' that shows all of the columns that contain
NA values
```{r}

a <- c()
for(i in 1:length(colnames(data))) {
  b <- any(is.na(data[,i]) == TRUE)
  a <- c(a, b)
}
missing <- data[,a]
```
Columns with missing data: first_gen_status, sex, City,  Zip_code, 
dependency_status, efc, has_children, hs_gpa, transfer_gpa, term_gpa
total_hr_earned, act_score



Calculates proportion of NA values for each column. This step is important
as the percentages are used to determine which columns contain too many NA
values and should thus not be imputed and not be used in any data analysis
```{r}
percNA <- c()
for(i in 1:ncol(missing)) {
TF <- (is.na(missing[,i]))
percNA <- c(percNA, length(TF[TF == TRUE])/length(TF))
}
names(percNA) <- colnames(missing)
percNA
```
columns with high percentage of NA values(above 50%): total_hrs_earned, and act_score
colomns to impute: first_gen_status (6), sex (7), dependency_status (17), efc(20), has_children(21)



removing columns with >53 levels as required by mice() (impute function)
```{r}
#mice.data <- select(data, -'city', -'zip_code', -'major_code', -'major_desc')
#data.imp <- mice(mice.data, method = 'rf')
#ximp <- complete(data.imp)
#any(is.na(ximp) == TRUE)
#ximp <- cbind(ximp, data$zip_code)
#colnames(ximp) <- c(colnames(ximp)[-29], 'zip_code')
#ximp <- na.omit(ximp)
```


The imputation process takes about 15 minutes to compute, so I exportes the 
imputed data set so that I dont have to re-impute everytime I worked in this
project
```{r}
#write.csv(ximp,'ximp.csv', row.names = FALSE)
```

importing imputed daaset
```{r}
ximp <- read.csv('/Users/gabby/Documents/Class Archive/Spring 2023/Senior Project/ximp.csv',
                 header = TRUE)
```


## Recoding
# ------------------------------------------------------------------------------


# Creating 'distance'' variable: calculates how far they live from campus

removing rows with invalid zip codes
```{r}
ximp <- ximp[is_zcta(ximp$zip_code) == TRUE,]
```
# calclulating distance from campus (campus zip = 80204)
```{r}
campzip <- rep(80204, nrow(ximp))
campdist <- zip_distance(ximp$zip_code, campzip)
ximp <- mutate(ximp, distance = campdist$distance)
```


# Converting Student type to Binary for regression
```{r}
ximp <- mutate(ximp,
               studenttype2 =
                 case_when(student_type == 'First-Time Freshmen' ~ 0,
                           student_type == 'Transfer Students' ~ 1
                           ))
#nrow(data[data$recode_student_type == 1,])/nrow(data)
# 54.99% transfer
```


# Converting Retention status to Binary for regression
```{r}
ximp <- mutate(ximp,
               retention2 =
                 case_when(retention_status == 'Not Retained' ~ 0,
                           retention_status == 'Retained' ~ 1))
#nrow(data[data$recode_retention == 1,])/nrow(data)
# 63.83 retention rate
```


# Converting Race to Binary for regression
```{r}
ximp <- mutate(.data = ximp,
               American_Indian_or_Alaskan_Native = ifelse(race_ethnicity == 
                                    'American Indian or Alaskan Native', 1, 0),
               Asian = ifelse(race_ethnicity == 'Asian' ,1, 0),
               Black_African_American = ifelse(race_ethnicity == 'Black or African American', 1, 0),
               Hispanic_Latino = ifelse(race_ethnicity == 'Hispanic or Latino', 1, 0),
               International = ifelse(race_ethnicity == 'International Students' , 1, 0),
               Native_Hawaiian_Pacific_Islander = ifelse(race_ethnicity == 'Native Hawaiian or Pacific Islander' , 1, 0),
               Two_or_More = ifelse(race_ethnicity == "Two or More Races" , 1, 0),
               White = ifelse(race_ethnicity == 'White', 1, 0))
ximp <- ximp[ximp$race_ethnicity != 'Other',]
 
```

# Converting first gen status to Binary for regression
```{r}
ximp <- mutate(ximp,
                firstgen2 = case_when(first_gen_status == 'First-Gen' ~ 1,
                 first_gen_status == 'Not First-Gen' ~ 0))

```


# Converting Sex to Binary for regression
```{r}

ximp <- mutate(ximp,
               sex2 = case_when(sex == 'F' ~ 1,
                                           sex == 'M' ~ 0))
```


# Converting Full / Part Time to Binary for regression
```{r}
ximp <- mutate(ximp,
               timestatus2 = case_when(time_status == 'Full-Time' ~ 1,
                                             time_status == 'Part-Time' ~ 0))
```

# Converting Dependency status to Binary for regression
```{r}
ximp <- mutate(ximp,
               depstatus2 = case_when(dependency_status == 'ind' ~ 0,
                                            dependency_status == 'dep' ~ 1))
```


# Converting Pell Eligable to Binary for regression
```{r}
ximp <- mutate(ximp,
               pell2= case_when(pell_eligible == 'Pell Eligible' ~ 1,
                                            pell_eligible == 'Not Pell Eligible' ~ 0))
```


# ConvertingOnline/ Hybrid/ In person to Binary for regression
```{r}
ximp <- mutate(ximp,
               inperson = ifelse(ximp$count_ftf > 0 | ximp$count_hyb > 0 , 1, 0)) 
```


# Converting has children to Binary for regression
```{r}
ximp <- mutate(ximp,
               child2 = ifelse(has_children == 'Y', 1, 0))
```


# Classification grade status based on GPA
```{r}
ximp <- mutate(ximp, grades = case_when(term_gpa < 2 ~ 'Poor',
                                       term_gpa >= 2 & term_gpa < 3 ~ 'Fair',
                                       term_gpa >= 3 ~ 'Good'))

```





#write.csv(ximp, 'grades_recoded.csv', row.names = F)

```

